---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# odr

<!-- badges: start -->
<!-- badges: end -->

The goal of *odr* is to help design efficient and well-powered multilevel experiments. We achieve this goal through three sets of functions. The *od* functions calculate the optimal sample size allocation by leveraging sampling cost information and other design parameters across levels and treatment conditions. The *re* function compares the relative efficiency across sample allocations. The *power* functions perform power analysis with and without accommodating cost and budget.

## Installation

You can install the released version of odr from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("odr")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(odr)
 # unconstrained optimal design
myod1 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
              varlim = c(0.01, 0.02))

 # constrained optimal design with n = 20
myod2 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
              plot.by = list(p = "p"), n = 20, varlim = c(0.005, 0.030))

 # relative efficiency
myre <- re(od = myod1, subod = myod2) # Functions rpe and re are same

# required budget calculation
mym <- power.2(expr = myod1, d = 0.3, q = 1, power = 0.8)

# See https://cran.r-project.org/web/packages/odr/vignettes/odr.html for more examples
```
