---
title: "Package 'odr'"
author: "Zuchao Shen & Benjamin Kelcey"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Package 'odr'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
The costs of sampling each additional unit in (multilevel) experimental studies
vary across treatment conditions (and levels of hierarchy). 
This package is a tool to identify the optimal sample allocation in experimental
studies such that statistical power is maximized under a fixed budget. 
The designs covers experiments  detecting main, mediation, and moderation
effects (and some of the combinations).

# 1. Optimal Design Parameters
Optimal design parameters in an optimal sample allocation include 

- the optimal sample sizes at each of the levels except the top level because 
the top-level sample size will be determined by total budget or power once all
other parameters are decided, and
- the proportion of units assigned to the treatment condition 
at the level of randomization.

Researchers can fix one or more of the optimal design parameters for 
constrained optimal sample allocation and solve for the remainder 
of optimal design parameters.


# 2. Designs Detecting Main Effects


## Designs and Functions

```{r, echo=FALSE}
df = data.frame(
  Design = c("Simple Experiments", "2-Level CRTs", "2-Level MRTs",
             "3-Level CRTs", "3-Level MRTs",
             "4-Level CRTs", "4-Level MRTs"),
  OD_Function = c("od.1", "od.2", "od.2m",
                  "od.3", "od.3m",
                  "od.4", "od.4m"),
  Power_Function = c("power.1", "power.2", "power.2m",
                     "power.3", "power.3m",
                     "power.4", "power.4m"),
  RE_Function = c("re", "re", "re",
                     "re", "re",
                     "re", "re")
  
)
names(df) <- c("Design", "OD Function", "Power Function", "RE Function")
library(knitr)
kable(df,
      caption = "Functions for Designs Detecting Main Effects")
```

## Rules for Function Names

For experiments detecting main effects, this package includes three categorical of functions and they are *od*, *power*, and *re*. For *od* and *power* functions, if a function name follows .# (e.g., od.2), this function is for #-level designs with treatment assignment at the top level (# in 1, 2, 3, 4). If a function name follows .#m (e.g., od.2m), this function is for #-level multisite-randomized trials.

## Functions

### *od* function

This function calculates optimal sample allocation parameters with and without 
constraint(s). For each type of multilevel experimental studies, there is an additional number (and one additional letter m) to be added to the general function name. For example, the function for two-level cluster randomized trials is *od.2*,
the function for two-level multisite randomized trials is *od.2m*.

### *power* function

This function performs power analyses with and without accommodating cost structures of sampling.
For power analysis accommodating cost structures, depending on which one parameter is left, the function calculates required budget (and sample size), statistical power, and minimum detectable effect size. For conventional power analysis without accommodating cost structures of sampling,
this function calculates required sample size, statistical power, and minimum detectable effect size. 
For each type of experiments, there is an additional number (and
an additional letter m) to be added to the general function name. For example, the function for 
two-level cluster randomized trials is *power.2*,
the function for two-level multisite randomized trials is *power.2m*.

### *re* function

This function calculates relative efficiency values between two designs. An alternative name for this function is *rpe* that stands for "relative precision and efficiency".


# 3. Designs Detecting Mediation Effects



## Designs and Functions

```{r, echo=FALSE}
df = data.frame(
  Design = c("Simple Experiments", "2-Level CRTs", "2-Level MRTs"),
  OD_Function = c("od.1.111", "od.2.221", "od.2m.111"),
  Power_Function = c("power.1.111", "power.2.221", "power.2m.111")
)
names(df) <- c("Design", "OD Function", "Power Function")
library(knitr)
kable(df,
      caption = "Functions for Designs Detecting Mediation Effects (In Preparation)")
```


## Rules for Function Names
There are two categories of functions for designs detecting mediation effects. They are *od* and *power* functions. The function names follow additional rules as these functions will add additional numbers to represent the levels of the treatment assignment, the mediator and the outcome. For example, *od.2.221* is the *od* function for two-level cluster-randomized trials (.2) with the treatment assignment and the mediator at the level two, and the outcome at the level 1 (.221). 
*od.2m.111* is the *od* function for two-level multisite-randomized trials (.2m) with the treatment assignment, the mediator, and the outcome at the level one (.111). 

## Functions
There are two categorical functions for designs detecting mediation effects and they are *od* and *power*.  *od* function can calculate optimal design parameters with and without a constraint(s). *power* function can perform power analysis. 

# 4. Examples for Main Effects
##  Function *od* 
Given cost structure (i.e., the costs of sampling each unit at 
different levels and treatment conditions), this function solves 
the optimal sample allocation with and without constraints. 

To solve the optimal sample allocation of a two-level cluster-randomized trial, 
we need the following information

- icc: intraclass correlation coefficient
- r12: the proportion of level-one outcome variance explained by covariates
- r22: the proportion of level-two outcome variance explained by covariates
- c1: the cost of sampling each additional level-one unit in the control condition 
- c2: the cost of sampling each additional level-two unit in the control condition 
- c1t: the cost of sampling each additional level-one unit in the experimental condition 
- c2t: the cost of sampling each additional level-two unit in the experimental condition
- m: a total fixed budget used to plot the variance curves, default value is the cost of sampling 60 level-two units across treatment conditions.

```{r}
library(odr)
```

```{r fig.width = 7, fig.height = 3.5}
 # unconstrained optimal design
myod1 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
              varlim = c(0.01, 0.02))
 # The function by default prints messages of output and plots the variance curves; one can turn off message and specify one or no plot.
 # myod1$out # output; 
 # myod1$par # parameters used in the calculation.
```

```{r fig.width = 5, fig.height = 5}
 # constrained optimal design with n = 20
myod2 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, 
              c2t = 50, plot.by = list(p = "p"), n = 20, 
              varlim = c(0.005, 0.030))
 # myod2$out # output
 # myod2$par # parameters used in the calculation.
```

```{r fig.width = 7, fig.height = 3.5}
 # constrained optimal design with p = 0.5
myod3 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50, 
             p = 0.5, varlim = c(0.005, 0.020))
 # myod3$out # output; 
 # myod3$par # parameters used in the calculation.
```

```{r fig.width = 7, fig.height = 3.5}
 # constrained n and p, no calculation performed
myod4 <- od.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
              plots = FALSE, n = 20, p = 0.5, varlim = c(0.005, 0.025))
```

Please see additional examples in corresponding functions by uncommenting below lines.
```{r}
# ?od.1 
# ?od.3
# ?od.4
# ?od.2m
# ?od.3m
# ?od.4m
```

## Function *power*
This function by default can perform power analyses accommodating cost structures 
(i.e., cost.model = TRUE), one of 'power', 'm', and 'd' must be NULL. For example,
if 'power' is NULL, the function calculates statistical power 
under a fixed budget and cost structure; if 'd' is NULL, the function
calculates minimum detectable effect size (i.e., d) under a fixed budget and 
desired power level; if 'm' is NULL, the function calculate
required budget (and required sample size) to achieve desired power level to detect a treatment effect.

This function also can conduct conventional power analysis or power analysis without 
accommodating cost structures by specifying cost.model = FALSE, the conventional
power analyses include statistical power calculation, minimum detectable effect size calculation, 
and required sample size calculation.

###  Examples of power analyses accommodating cost structures (cost.model = TRUE)
###### Required budget for desired power
- Required budget calculation
```{r}
mym <- power.2(expr = myod1, d = 0.3, q = 1, power = 0.8)
# mym$out  # m =1702, J = 59
```
- Effects on required budget to maintain same level power when designs depart from the optimal one 
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c(1, 2))
budget <- NULL
nrange <- c(2:50)
for (n in nrange)
  budget <- c(budget, power.2(expr = myod1, constraint = list (n = n), d = 0.3, q = 1, power = 0.8)$out$m)
plot(nrange, budget, type = "l", lty = 1, xlim = c(0, 50), ylim = c(1500, 3500),
     xlab = "Level-1 sample size: n", ylab = "Budget", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
budget <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  budget <- c(budget, power.2(expr = myod1, constraint = list (p = p), d = 0.3, q = 1, power = 0.8)$out$m)
plot(prange, budget, type = "l", lty = 1, xlim = c(0, 1), ylim = c(1500, 7000),
     xlab = "Porportion groups in treatment: p", ylab = "Budget", main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
par(figure)
```

###### Statistical power under a fixed budget
- Power calculation
```{r}
mypower <- power.2(expr = myod1, q = 1, d = 0.3, m = 1702)
# mypower$out  # power = 0.80
```
- Effects on power under same budget when designs depart from the optimal one
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
pwr <- NULL
nrange <- c(2:50)
for (n in nrange)
  pwr <- c(pwr, power.2(expr = myod1, constraint = list (n = n), d = 0.3, q = 1, m = 1702)$out)
plot(nrange, pwr, type = "l", lty = 1, xlim = c(0, 50), ylim = c(0.4, 0.9),
     xlab = "Level-1 sample size: n", ylab = "Power", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
pwr <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  pwr <- c(pwr, power.2(expr = myod1, constraint = list (p = p), d = 0.3, q = 1, m = 1702)$out)
plot(prange, pwr, type = "l", lty = 1, xlim = c(0, 1), ylim = c(0.1, 0.9),
     xlab = "Porportion groups in treatment: p", ylab = "Power",  main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
 par(figure)
```

###### Minimum detectable effect size under a fixed budget
- minimum detectable effect size calculation
```{r}
mymdes <- power.2(expr = myod1, q = 1, power = 0.80, m = 1702)
# above experssion takes parameters and outputs from od.2 function. Equivalently, each parameter can be explicitly specified.
# mym <- power.2(icc = 0.2, r12 = 0.5, r22 = 0.5, c1 = 1, c2 = 5, c1t = 1, c2t = 50,
#                     n = 9, p = 0.33, d = 0.3, q = 1, power = 0.8)
# mymdes$out  # d = 0.30
```
- Effects on minimum detectable effect size under same budget when designs depart from the optimal one 
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
MDES <- NULL
nrange <- c(2:50)
for (n in nrange)
  MDES <- c(MDES, power.2(expr = myod1, constraint = list (n = n), power = 0.8, q = 1, m = 1702)$out)
plot(nrange, MDES, type = "l", lty = 1, xlim = c(0, 50), ylim = c(0.3, 0.8),
     xlab = "Level-1 sample size: n", ylab = "MDES", main = "", col = "black")
 abline(v = 9, lty = 2, col = "Blue")
 
MDES <- NULL
prange <- seq(0.05, 0.95, by = 0.005)
for (p in prange)
  MDES <- c(MDES, power.2(expr = myod1, constraint = list (p = p), power = 0.8, q = 1, m = 1702)$out)
plot(prange, MDES, type = "l", lty = 1, xlim = c(0, 1), ylim = c(0.3, 0.8),
     xlab = "Porportion groups in treatment: p", ylab = "MDES", main = "", col = "black")
 abline(v = 0.33, lty = 2, col = "Blue")
 par(figure)
```

### Examples of conventional power analyses (cost.model = FALSE)
```{r}
# Required level-2 sample size calculation
myJ <- power.2(cost.model = FALSE, expr = myod1, d = 0.3, q = 1, power = 0.8)
# above experssion takes parameters and outputs from od.2 function. Equivalently, each parameter can be explicitly specified.
# myJ <- power.2(icc = 0.2, r12 = 0.5, r22 = 0.5, 
#                     cost.model = FALSE, n = 9, p = 0.33, d = 0.3, q = 1, power = 0.8)
myJ$out  # J = 59

# Power calculation
mypower1 <- power.2(cost.model = FALSE, expr = myod1, J = 59, d = 0.3, q = 1)
mypower1$out  # power = 0.80

# Minimum detectable effect size calculation
mymdes1 <- power.2(cost.model = FALSE, expr = myod1, J = 59, power = 0.8, q = 1)
mymdes1$out  # d = 0.30
```
###  Examples of conventional power curves
```{r fig.width = 7, fig.height = 3.5}
figure <- par(mfrow = c (1, 2))
pwr <- NULL
mrange <- c(300:3000)
for (m in mrange)
  pwr <- c(pwr, power.2(expr = myod1, d = 0.3, q = 1, m = m)$out)
plot(mrange, pwr, type = "l", lty = 1, xlim = c(300, 3000), ylim = c(0, 1),
     xlab = "Budget", ylab = "Power", main = "", col = "black")
 abline(v = 1702, lty = 2, col = "Blue")
 
pwr <- NULL
Jrange <- c(4:100)
for (J in Jrange)
  pwr <- c(pwr, power.2(expr = myod1, cost.model = FALSE, d = 0.3, q = 1, J = J)$out)
plot(Jrange, pwr, type = "l", lty = 1, xlim = c(4, 100), ylim = c(0, 1),
     xlab = "Level-2 sample size: J", ylab = "Power", main = "", col = "black")
 abline(v = 59, lty = 2, col = "Blue")
par(figure)
```

###  Examples for other types of trials
Please see examples in corresponding functions by uncommenting below lines.
```{r}
# ?power.1
# ?power.3
# ?power.4
# ?power.2m
# ?power.3m
# ?power.4m
```

##  Function *re* 
Calculate the relative efficiency (RE) of two designs, this function uses the returns from *od* function

Based on above examples in *od* functions, calculate the relative efficiency
```{r}
# relative efficiency (RE) of a constrained design comparing with the optimal design
myre <- re(od = myod1, subod= myod2)
myre$re # get the output (i.e., RE = 0.88)

# relative efficiency (RE) of a constrained design comparing with the unconstrained optimal one
myre <- re(od = myod1, subod= myod3)

# relative efficiency (RE) of a constrained design comparing with the unconstrained optimal one
myre <- re(od = myod1, subod= myod4)


```

For additional examples, please see example sections in corresponding *od* functions by uncommenting below lines.
```{r}
# ?od.1
# ?od.2
# ?od.3
# ?od.4
# ?od.2m
# ?od.3m
# ?od.4m
```

# 5. Examples for Mediation Effects
Below is a simple example to calculate optimal sample allocation and perform statistical power analyses.

```{r}
# Optimal sample allocation and statistical power for randomized controlled trials
myod <- od.1.111(a = .3, b = .5, c1 = 10, c1t = 100, verbose = FALSE)
mypower <- power.1.111(expr = myod, power = .8)
# mypower

# Conventional power analyses
mypower <- power.1.111(cost.model = FALSE, a = .3, b = .5, test = "joint",
                        power = .8, p =.5)
# mypower
```
